Type theory in type theory using single substitutions

### 2024-05-07 types_2024/abstract.tex
typo
final

In this abstract we define type theory in a minimalistic way: we aim
for an intrinsic, that is, well-typed and quotiented definition of the
syntax, that is as small as possible in terms of number of operations
and equations for the substitution calculus. We only introduce
operations that are forced upon us. We aim for a language with $\Pi$
types, and it turns out that we have to add all the ingredients of the
substitution calculus while introducing the usual rules for $\Pi$
types (abstraction, application, $\beta$ and $\eta$). When adding more
type formers, we don't need to extend the substitution calculus
anymore.

@@

types and variables.

@@

substitution $\p$. For now, $\Sub$ seems like an overkill because we

@@

possible substitution: weakening of a variable increases the index by

@@

Finally, to typecheck the $\Pi\eta$ rule, we need our last equation on

@@

This concludes all the rules for type theory with $\Pi$. We summarise

@@

substitution, parallel weakenings, or other combinations of these.

When adding new type formers, we only need the rules for the type
former, and an extra instantiation (naturality) rule for each
operation. For example, a Coquand-universe can be added by $\U :
\Ty\,\Gamma$, $\El:\Tm\,\Gamma\,\U\ra\Ty\,\Gamma$,
$\c:\Ty\,\Gamma\ra\Tm\,\Gamma\,\U$, $\U\beta : \El\,(\c\,A) = A$,
$\U\eta : \c\,(\El\,a) = a$, and three instantiation rules (note that
we need indexing to avoid inconsistency). In \cite{sogat}, we showed
that any second-order generalised algebraic theory (SOGAT) has a
single substitution presentation.

In the syntax (initial model, quotient inductive-inductive type) of
the above theory, all the rules of categories with families (CwF
\cite{DBLP:journals/corr/abs-1904-00827}) are admissible. That is, by
induction on the single substitution syntax, we can define parallel
substitutions (lists of terms) which are composable and form a
category; we can define instantiation by parallel substitutions for
types and terms, these have the usual universal property of
comprehension. The main ingredient for this construction is the notion
of $\alpha$-normal form: a kind of normal form which is still
quotiented by $\Pi\beta$, $\Pi\eta$, but does not include explicit
substitutions. If a type is in $\alpha$-normal form, we know whether
it is $\Pi$, $\U$, or $\El$ of a term. If a term is in $\alpha$-normal
form, we know whether it is a variable, a $\lam$, an application or a
code for a type (note that any function on an $\alpha$-normal has to
respect $\Pi\beta$, $\Pi\eta$, $\U\beta$, $\U\eta$). We prove
$\alpha$-normalisation (every term has a unique $\alpha$-normal form)
and then use induction on $\alpha$-normal forms to define parallel
instantiation and prove its properties.

We formalised single substitution calculus with an infinite hierarchy
of types closed under $\Pi$ and $\U$ as a quotient inductive-inductive
type in Agda, proved $\alpha$-normalisation, and derived all the rules
for parallel substitutions (CwF equipped with $\Pi$ types and $\U$):
\url{https://bitbucket.org/akaposi/single}.

It is clear that the rules for the single substitution calculus are
all derivable from the CwF-rules. The other direction is however not
true: there are more models of the single substitution calculus than
the parallel substitution calculus. The situation is analogous to the
relationship of combinatory logic and lambda calculus: the former has
more models, but the syntaxes are isomorphic
\cite{DBLP:conf/fscd/AltenkirchKSV23}. Our single substitution
calculus is a minimalistic version of B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023} (C-systems are their
parallel substitution analogue): B-systems include telescopes and more
equations than our single substitution calculus, making them
equivalent to C-systems.

>>>>

What is the syntax of Martin-Löf type theory? In this abstract, we
give a minimalistic answer to this question. We aim to define the
syntax of type theory only using operations that are unavoidable. We
would also like to eschew boilerplate by only defining well-typed terms
which are quotiented by conversion
\cite{DBLP:conf/popl/AltenkirchK16}. This means that type theory is a
generalised algebraic theory (GAT), its syntax is a quotient
inductive-inductive type (QIIT). Usual such definitions of type theory are
based on categories: category with families \cite{DBLP:journals/corr/abs-1904-00827}, locally cartesian closed
category \cite{Seely1984-SEELCC,DBLP:journals/mscs/ClairambaultD14}, contextual category \cite{DBLP:journals/apal/Cartmell86,initiality-agda}, C systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}, etc. In this abstract we
show that categories are not necessary for defining the GAT of type
theory. A category-free definition of type theory is B-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023} which
includes telescopes with complex operations and equations. In our
definition we avoid these as well. In this abstract we give a
tutorial-style introduction to our minimalistic definition of type
theory, no prior knowledge of the metatheory of type theory is
required. We introduce the operations in a naive, logical order.

@@

types (which depend on a context) and variables (which are in a context and have a type).

@@

substitution $\p$. For now, a separate sort $\Sub$ seems like an overkill because we

@@

possible substitution $\Sub$: weakening of a variable increases the index by

@@

Finally, to typecheck the $\Pi\eta$ rule, we need our last equations on

@@

This concludes all the rules for type theory with $\Pi$ (this type theory is actually empty because there are no base types, but is enough to illustrate our method). We summarise

@@

substitution, parallel weakenings, telescopes, or combinations of these.

When adding new type formers, we only need the rules for the type
former, and an extra instantiation (naturality) rule for each
operation. For example, a Coquand-universe can be added by $\U :
\Ty\,\Gamma$, $\El:\Tm\,\Gamma\,\U\ra\Ty\,\Gamma$,
$\c:\Ty\,\Gamma\ra\Tm\,\Gamma\,\U$, $\U\beta : \El\,(\c\,A) = A$,
$\U\eta : \c\,(\El\,a) = a$, and three instantiation rules (note that
we would need indexing $\U$ and $\Ty$ by universe levels to avoid inconsistency). In \cite{sogat}, we showed
that any second-order generalised algebraic theory has a
single substitution presentation.

In the syntax (initial model, QIIT) of
the above theory, all the rules of categories with families (CwF
\cite{DBLP:journals/corr/abs-1904-00827}) are admissible. That is, by
induction on the single substitution syntax, we can define parallel
substitutions (lists of terms) which are composable and form a
category; we can define instantiation by parallel substitutions for
types and terms, these have the usual universal property of
comprehension. The main ingredient for this construction is the notion
of $\alpha$-normal form: a kind of normal form which is still
quotiented by $\Pi\beta$, $\Pi\eta$, but does not include explicit
substitutions. If a type is in $\alpha$-normal form, we know whether
it is $\Pi$, $\U$, or $\El$ of a term. If a term is in $\alpha$-normal
form, we know whether it is a variable, a $\lam$, an application or a
code for a type (note that any function on an $\alpha$-normal type/term has to
respect $\Pi\beta$, $\Pi\eta$, $\U\beta$, $\U\eta$). We prove
$\alpha$-normalisation (every term has a unique $\alpha$-normal form)
and then use induction on $\alpha$-normal forms to define parallel
instantiation and prove its properties.

We formalised a single substitution calculus with an infinite hierarchy
of types closed under $\Pi$ and $\U$ as a QIIT in Agda, proved
$\alpha$-normalisation, and derived all the rules for parallel
substitutions (CwF equipped with $\Pi$ types and $\U$):
\url{https://bitbucket.org/akaposi/single}. In the formalisation, we
use $\mathsf{sProp}$-valued equality and the syntax is postulated as a
QIIT with rewrite rules for its $\beta$ laws.

It is clear that the rules for the single substitution calculus are
all derivable from the CwF-rules. The other direction is however not
true: there are more models of the single substitution calculus than
the parallel substitution calculus, but the syntaxes are
isomorphic. The situation is analogous to the relationship of
combinatory logic and lambda calculus: their syntaxes are isomorphic,
but the former has more models \cite{DBLP:conf/fscd/AltenkirchKSV23}.
